---
title: "Bootstrap Inference for Logistic Regression"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bootstrap Inference for Logistic Regression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi       = 150,
  out.width = "100%"
)
set.seed(42)
```

## Overview

**strapwise** provides two functions:

- `bootstrap_logistic()` — applies case-resampling bootstrap to a fitted
  `glm` object with `family = binomial`, producing a matrix of bootstrapped
  coefficients and percentile confidence intervals.
- `plot_logistic_curves()` — generates one marginal response-curve plot per
  predictor: a ribbon + probability curve for continuous predictors, and a
  dot + error bar chart for categorical predictors, with observed binary
  outcomes overlaid as jittered points on both.

---

## Simulated data

We simulate a binary outcome influenced by age (continuous), BMI
(continuous), sex (two-level factor), and treatment (three-level ordered
factor).

```{r simulate}
library(strapwise)

set.seed(42)
n  <- 400
df <- data.frame(
  age = rnorm(n, mean = 55, sd = 12),
  bmi = rnorm(n, mean = 27, sd =  5),
  sex = factor(sample(c("Female", "Male"), n, replace = TRUE)),
  trt = factor(
    sample(c("Placebo", "Low", "High"), n, replace = TRUE,
           prob = c(0.33, 0.33, 0.34)),
    levels = c("Placebo", "Low", "High")
  )
)

df$y <- rbinom(n, 1, plogis(
  -3 +
  0.04 * df$age +
  0.06 * df$bmi +
  0.50 * (df$sex == "Male") +
  0.80 * (df$trt == "Low") +
  1.50 * (df$trt == "High")
))

table(df$y)
```

---

## Fit the logistic regression model

Model fitting is the user's responsibility. `strapwise` works with any
`glm(..., family = binomial)` object—the link function (logit, probit,
complementary log-log, etc.) is detected automatically.

```{r fit}
fit <- glm(y ~ age + bmi + sex + trt, data = df, family = binomial)
summary(fit)
```

---

## Run the bootstrap

`bootstrap_logistic()` resamples the original data with replacement and
refits the model `n_boot` times. Replicates that error or generate warnings
during refitting are skipped; if any are skipped, the function reports how
many.

```{r bootstrap}
bl <- bootstrap_logistic(fit, n_boot = 1000, seed = 42, conf_level = 0.95)
print(bl)
```

`summary()` returns a data frame of original-data estimates with bootstrapped
standard errors, bias, and percentile confidence intervals.

```{r summary}
summary(bl)
```

---

## Response curve plots

`plot_logistic_curves()` returns a named list of `ggplot` objects—one per
predictor. Non-focal predictors are held at their **median** (numeric) or
**mode** (factor).

### Continuous predictors

Each continuous-predictor plot shows:

- A probability curve (original-data model fit).
- A shaded ribbon (95% bootstrap CI).
- Jittered raw 0/1 outcomes at their true y-value, providing direct
  visualisation of the data density and balance.

```{r plot-age, fig.cap = "Marginal response curve for age."}
plots <- plot_logistic_curves(bl)
plots[["age"]]
```

```{r plot-bmi, fig.cap = "Marginal response curve for BMI."}
plots[["bmi"]]
```

### Categorical predictors

Each categorical-predictor plot shows:

- A predicted-probability marker per level (filled dot).
- 95% bootstrap CI error bars.
- Jittered 0/1 raw outcomes within each category strip, coloured by level.

```{r plot-sex, fig.cap = "Predicted probability of response by sex."}
plots[["sex"]]
```

```{r plot-trt, fig.cap = "Predicted probability of response by treatment group."}
plots[["trt"]]
```

---

## Combined figure

If **patchwork** is installed all four plots can be assembled in one call.

```{r combine, fig.width = 12, fig.height = 8, fig.cap = "All response curves."}
# install.packages("patchwork")   # if needed
plot_logistic_curves(bl, combine = TRUE)
```

---

## Alternative link functions

The functions work unchanged with any `binomial` link. The `linkinv`
function is read directly from the fitted model object, so back-transformation
is always consistent with the link used.

```{r probit}
fit_probit <- glm(y ~ age + bmi + sex + trt, data = df,
                  family = binomial(link = "probit"))
bl_probit  <- bootstrap_logistic(fit_probit, n_boot = 500, seed = 1)
plot_logistic_curves(bl_probit)[["age"]]
```

---

## Comparing models with ROC curves

`compare_logistic_roc()` overlays ROC curves from multiple fitted binomial
GLMs in a single `ggplot`, making it easy to compare discrimination across
candidate models.

```{r roc-comparison, fig.cap = "Overlaid ROC curves for two logistic regression models."}
fit_age <- glm(y ~ age, data = df, family = binomial)
fit_full <- glm(y ~ age + bmi + sex + trt, data = df, family = binomial)

compare_logistic_roc(
  list("Age only" = fit_age, "Full model" = fit_full)
)
```

---

## Inline formula transformations

Formulae with inline transformations such as `log()` or `I()` are supported.
`bootstrap_logistic()` stores the original untransformed data and lets
`model.matrix()` apply the transformation during each bootstrap refit.
The resulting plot uses the base variable name on the x-axis, with
predictions made on the transformed scale.

```{r log-transform}
fit_log <- glm(y ~ log(age) + bmi + sex, data = df, family = binomial)
bl_log  <- bootstrap_logistic(fit_log, n_boot = 500, seed = 1)
plot_logistic_curves(bl_log)[["age"]]
```

---

## Customising plots

All visual parameters are exposed as arguments to `plot_logistic_curves()`.
Plots are standard `ggplot` objects and can be further modified after the
fact.

```{r custom, fig.cap = "Customised response curve for age."}
library(ggplot2)

plot_logistic_curves(
  bl,
  palette        = c("#8B1A1A", "#2F4F8B"),
  ribbon_alpha   = 0.15,
  raw_alpha      = 0.15,
  raw_point_size = 1,
  jitter_height  = 0.015,
  base_theme     = theme_minimal(base_size = 13)
)[["age"]] +
  labs(title = "Custom theme: age response curve",
       caption = "Shading = 95% bootstrap CI")
```
