---
title: "Forest Plots for Logistic and Cox Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Forest Plots for Logistic and Cox Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 10,
  fig.height = 5,
  dpi        = 150,
  out.width  = "100%"
)
set.seed(42)
```

## Overview

**strapwise** provides two functions for building publication-ready forest
plots from fitted regression models:

- `fp_data()` — extracts coefficients and confidence intervals from a `glm`
  or `coxph` object and returns a structured data frame suitable for
  inspection or modification before plotting.
- `fp_plot()` — calls `fp_data()` internally and renders the forest plot in
  one step.  All `fp_data()` arguments are forwarded directly.

Both functions auto-detect the model type from the object class and produce
the appropriate column header (`OR (95% CI)` for logistic models,
`HR (95% CI)` for Cox models).

```{r load}
library(strapwise)
library(survival)
```

---

## Part 1 — Logistic regression

### Data

We simulate 500 observations with three predictors:

- `age_log` — patient age stored on the log scale (a common transformation
  when the raw distribution is right-skewed).
- `bmi` — body-mass index on the natural scale.
- `ecog` — a binary ECOG performance-status factor with levels `"0"` and
  `"1+"`.

```{r logistic-data}
n       <- 500L
age     <- rnorm(n, mean = 60, sd = 12)
age_log <- log(pmax(age, 1))
bmi     <- rnorm(n, mean = 27, sd = 5)
ecog    <- factor(sample(c("0", "1+"), n, replace = TRUE,
                          prob = c(0.6, 0.4)))

lp <- -6 + 1.8 * age_log + 0.06 * bmi + 0.8 * (ecog == "1+")
y  <- rbinom(n, 1L, plogis(lp))

logistic_dat <- data.frame(y, age_log, bmi, ecog)
```

### Model

```{r logistic-fit}
logistic_fit <- glm(
  y ~ age_log + bmi + ecog,
  data   = logistic_dat,
  family = binomial
)
summary(logistic_fit)
```

### Inspecting the intermediate data frame

`fp_data()` is useful when you need to examine or post-process the plot
data before rendering.  Two arguments control how predictor values are
displayed:

- **`labels`** — a named character vector mapping raw variable names to
  publication-ready display labels.  Unmatched names fall back to the raw
  name.
- **`log_vars`** — a character vector of variable names that are stored on
  the log scale.  The percentile display values for these variables are
  back-transformed with `exp()` so readers see the original units (e.g.
  years, not log-years).

```{r logistic-fp-data}
logistic_labels <- c(
  age_log = "Age (years)",
  bmi     = "BMI (kg/m²)",
  ecog    = "Baseline ECOG PS"
)

logistic_pd <- fp_data(
  model    = logistic_fit,
  data     = logistic_dat,
  labels   = logistic_labels,
  log_vars = "age_log",
  percs    = c(0.1, 0.25, 0.75, 0.9)
)

print(logistic_pd[, c("Predictor", "OR (95% CI)", "P-value", "is_header")])
```

Each predictor occupies one *group*: a bold header row (`is_header = TRUE`)
followed by sub-rows for factor levels or percentile values.

### Rendering the plot

`fp_plot()` accepts the same arguments as `fp_data()` plus plot-specific
controls, so a single call builds and renders the plot:

```{r logistic-plot}
fp_plot(
  logistic_fit,
  logistic_dat,
  labels   = logistic_labels,
  log_vars = "age_log",
  percs    = c(0.1, 0.25, 0.75, 0.9)
)
```

### Customisation

```{r logistic-custom, fig.height = 5}
# Tighter x limits and labelled tick marks
fp_plot(
  logistic_fit, logistic_dat,
  labels   = logistic_labels,
  log_vars = "age_log",
  percs    = c(0.1, 0.25, 0.75, 0.9),
  xlim     = c(0.5, 5),
  ticks_at = c(0.5, 1, 2, 5)
)
```

```{r logistic-linear, fig.height = 5}
# Linear x-axis — useful when all ORs are close to 1
fp_plot(
  logistic_fit, logistic_dat,
  labels   = logistic_labels,
  log_vars = "age_log",
  percs    = c(0.1, 0.25, 0.75, 0.9),
  x_trans  = "none",
  xlim     = c(0, 4)
)
```

---

## Part 2 — Cox proportional hazards

### Data

We use the `lung` dataset from the **survival** package.  `meal_log` stores
meal calories on the log scale, so we pass it to `log_vars` to recover
original units in the percentile labels.

```{r cox-data}
lung <- survival::lung
lung$status   <- lung$status - 1L   # 0 = censored, 1 = event
lung$sex      <- factor(lung$sex, levels = 1:2,
                         labels = c("Male", "Female"))
lung$ecog_cat <- factor(
  ifelse(lung$ph.ecog >= 2L, "2+", as.character(lung$ph.ecog)),
  levels = c("0", "1", "2+")
)
lung$meal_log <- log(pmax(lung$meal.cal, 1))

cox_dat <- lung[complete.cases(
  lung[, c("time", "status", "age", "sex", "ecog_cat", "meal_log")]
), ]

nrow(cox_dat)
```

### Model

```{r cox-fit}
cox_fit <- coxph(
  Surv(time, status) ~ age + sex + ecog_cat + meal_log,
  data = cox_dat
)
summary(cox_fit)
```

### Forest plot

`fp_plot()` detects the `coxph` class and uses `HR (95% CI)` as the column
header and `"Lower / Greater hazard"` as arrow labels automatically.

```{r cox-plot}
cox_labels <- c(
  age      = "Age (years)",
  sex      = "Sex",
  ecog_cat = "Baseline ECOG PS",
  meal_log = "Meal calories (kcal)"
)

fp_plot(
  cox_fit, cox_dat,
  labels   = cox_labels,
  log_vars = "meal_log",
  percs    = c(0.1, 0.25, 0.75, 0.9)
)
```

### Customisation

```{r cox-custom, fig.height = 5}
# Narrower x range; suppress margin lines; alternative colour
fp_plot(
  cox_fit, cox_dat,
  labels    = cox_labels,
  log_vars  = "meal_log",
  percs     = c(0.1, 0.25, 0.75, 0.9),
  xlim      = c(0.3, 3),
  ticks_at  = c(0.3, 0.5, 1, 2, 3),
  vert_line = NULL,
  ci_col    = "#1A9641"
)
```

---

## Saving to file

`fp_plot()` returns a `gtable` object.  Use `grid::grid.draw()` inside a
graphics device to save it:

```{r save, eval = FALSE}
fp <- fp_plot(
  cox_fit, cox_dat,
  labels   = cox_labels,
  log_vars = "meal_log"
)

png("cox_forest.png", width = 10, height = 5, units = "in", res = 300)
grid::grid.draw(fp)
dev.off()
```

---

## Interaction terms

`fp_plot()` / `fp_data()` handle interaction terms automatically:

- A **continuous × continuous** interaction (e.g. `age:bmi`) is shown as a
  single row carrying the OR/HR for a 1-unit change in the product term.
- A **continuous × factor** interaction (e.g. `age:sex`) is shown as a
  header row with one child row per non-reference factor level.

```{r interaction}
set.seed(7L)
n_int   <- 400L
age_i   <- rnorm(n_int, 55, 10)
sex_i   <- factor(sample(c("F", "M"), n_int, replace = TRUE))
y_int   <- rbinom(n_int, 1L,
                  plogis(-4 + 0.05 * age_i + 0.5 * (sex_i == "M") +
                           0.01 * age_i * (sex_i == "M")))
int_df  <- data.frame(age = age_i, sex = sex_i, y = y_int)

int_fit <- glm(y ~ age + sex + age:sex, data = int_df, family = binomial)

fp_plot(
  int_fit, int_df,
  labels = c(age = "Age (years)", sex = "Sex",
             "age:sex" = "Age × Sex interaction")
)
```
